pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-14 manuscript/greenwell.tex
minor edits

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression). Unfortunately, it can be challenging for the uninitiated to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

@@

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details. PDPs help visualize the relationship between a subset of the predictors (typically 1-3) and the response while accounting for the average effect of the other independent variables in the model. They are particularly effective with black box models like random forests and support vector machines.

@@

\begin{table}[htbp]
  \begin{tabular}{p{4cm}ll}
    \toprule
      Type of model & R package & Object class \\
      \midrule
      Decision tree             & \CRANpkg{C50} \citep{C50-pkg} & \code{"C5.0"} \\
                                & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
                                & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
      Bagged decision trees     & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
                                & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
      Boosted decision trees    & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
                                & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
      Cubist                    & \CRANpkg{Cubist} \citep{Cubist-pkg} & \code{"cubist"} \\
      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
      Support vector machine    & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                                & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{pdp} package. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} for details.}
  \label{tab:models}
\end{table}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_rf_vimp}
  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
  \label{fig:plotmo_vs_partial}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
  \label{fig:pd_lstat}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_extrap}
  \caption{Examples of PDPs with the addition of a rug display (left) and a convex hull (right).}
  \label{fig:partial_extrap}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_par}
  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}. Since \code{dpg} is contiuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
  \label{fig:partial_par}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
  \caption{Partial dependence of species on petal width and petal length for the iris data.}
  \label{fig:partial_iris}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_xgb}
  \caption{PDPs for the top two most important variables in the Boston housing data using XGBoost. Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
  \label{fig:boston_xgb}
\end{figure}

>>>>

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

@@

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details. PDPs help visualize the relationship between a subset of the features (typically 1-3) and the response while accounting for the average effect of the other predictors in the model. They are particularly effective with black box models like random forests and support vector machines.

@@

\begin{table}[!htbp]
  \begin{tabular}{p{4cm}ll}
    \toprule
      Type of model & R package & Object class \\
      \midrule
      Decision tree             & \CRANpkg{C50} \citep{C50-pkg} & \code{"C5.0"} \\
                                & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
                                & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
      Bagged decision trees     & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
                                & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
      Boosted decision trees    & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
                                & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
      Cubist                    & \CRANpkg{Cubist} \citep{Cubist-pkg} & \code{"cubist"} \\
      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
      Support vector machine    & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                                & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{pdp} package. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} for details.}
  \label{tab:models}
\end{table}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_rf_vimp}
  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
  \label{fig:plotmo_vs_partial}
\end{figure}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
  \label{fig:pd_lstat}
\end{figure}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_extrap}
  \caption{Examples of PDPs with the addition of a rug display (left) and a convex hull (right).}
  \label{fig:partial_extrap}
\end{figure}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_par}
  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}. Since \code{dpg} is contiuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
  \label{fig:partial_par}
\end{figure}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
  \caption{Partial dependence of species on petal width and petal length for the iris data.}
  \label{fig:partial_iris}
\end{figure}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_xgb}
  \caption{PDPs for the top two most important variables in the Boston housing data using XGBoost. Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
  \label{fig:boston_xgb}
\end{figure}
